---
title: "L.V. Causality Marketing Assignment 2"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    self-contained-math: true
toc: true
---

```{r}
#| label: load necessary packages
#| include: false

library(sjPlot)
library(kableExtra)
library(ggplot2)
library(AER)
library(MatchIt)
library(sandwich)
library(lmtest)
library(DoubleML)
library(mlr3)
library(mlr3learners)
library(glmnet)
library(ranger)
library(xgboost)
```

# Import Data
```{r}
casedata_task3 <- read.csv("CaseData_Task3.csv", stringsAsFactors = FALSE)

# Display the structure of the data to verify column names
str(casedata_task3)

# Check the first few rows
head(casedata_task3)
```
<!-- # Use only treatment + appropriate pre-treatment covariates -->

# Estimate the treatment effect using simple linear Regression Adjustment
The estimate for the discount coefficient is the Regression Adjustment ATE
```{r}
model_adj <- lm(y ~ d + X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 +
                 X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20,
                 data = casedata_task3)

summary(model_adj)
```

## Specifically extract the ATE value
```{r}
ate <- coef(model_adj)["d"]
cat("The estimated Average Treatment Effect (ATE) for the treatment is:", ate, "\n")
```

```{r}
# Optional: robust SE
coeftest(model_adj, vcov = vcovHC(model_adj, type = "HC3"))
```

# Matching comparisons
Preparing the data for matching excersises
```{r}
ps_formula <- d ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20
```

## 1:1 Nearest Neighbor Matching WITHOUT Replacement
```{r}
cat("\n--- Matching: 1:1 Without Replacement ---\n")
match_no_replace <- matchit(ps_formula, 
                            data = casedata_task3, 
                            method = "nearest", 
                            distance = "glm",   # Uses logistic regression for propensity scores
                            replace = FALSE,    # Without replacement
                            ratio = 1)          # 1:1 matching

# View a summary to check covariate balance (ideally Std. Mean Diff < 0.1)
cat("Balance Summary (Without Replacement):\n")
print(summary(match_no_replace, un = FALSE))

# Extract the matched dataset
data_matched_no_replace <- match.data(match_no_replace)

# Estimate the Treatment Effect
# We include the covariates in the final outcome model to make it "doubly robust"
model_no_replace <- lm(y ~ d + X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20, 
                       data = data_matched_no_replace, 
                       weights = weights)

# Calculate robust standard errors
cat("\nTreatment Effect (Without Replacement):\n")
print(coeftest(model_no_replace, vcov. = vcovHC))
```

```{r}
att <- coef(model_no_replace)["d"]
cat("The estimated Average Treatment Effect on the Treated for the treatment is (ATT):", att, "\n")
```

## 1:1 Nearest Neighbor Matching WITH Replacement
```{r}
cat("\n--- Matching: 1:1 With Replacement ---\n")
match_with_replace <- matchit(ps_formula, 
                              data = casedata_task3, 
                              method = "nearest", 
                              distance = "glm",
                              replace = TRUE,     # With replacement
                              ratio = 1)

# View a summary to check covariate balance
cat("Balance Summary (With Replacement):\n")
print(summary(match_with_replace, un = FALSE))

# Extract the matched dataset
# Use get_matches() instead of match.data() when matching WITH replacement 
# to properly duplicate rows that are used as matches multiple times.
data_matched_with_replace <- get_matches(match_with_replace)

# Estimate the Treatment Effect
model_with_replace <- lm(y ~ d + X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20, 
                         data = data_matched_with_replace, 
                         weights = weights)

# For matching WITH replacement, individuals can be matched multiple times. 
# Therefore, standard errors must be clustered by the matched pair ID (`subclass`).
cat("\nTreatment Effect (With Replacement):\n")
print(coeftest(model_with_replace, vcov. = vcovCL, cluster = ~subclass))
```

```{r}
att <- coef(model_with_replace)["d"]
cat("The estimated Average Treatment Effect on the Treated for the treatment is (ATT):", att, "\n")
```




## Euclidean Distance Matching
```{r}
euc_out <- matchit(
  d ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 +
       X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20,
  data = casedata_task3,
  method = "nearest",
  distance = "euclidean",
  normalize = TRUE
)

summary(euc_out)
# bal.tab(euc_out) - balance info already shown in summary above

m_data_euc <- match.data(euc_out)
lm_euc <- lm(y ~ d + X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20, 
             data = m_data_euc, weights = weights)

cat("\nTreatment Effect (Euclidean Matching):\n")
summary(lm_euc)

cat("\nRobust Standard Errors (Euclidean Matching):\n")
print(coeftest(lm_euc, vcov. = vcovHC(lm_euc, type = "HC3")))
```


## Inverse Probability Weighting (IPW)
```{r}
# Complete-case sample for IPW variables
ipw_vars <- c("y", "d", paste0("X", 1:20))
df_ipw <- casedata_task3[complete.cases(casedata_task3[, ipw_vars]), ]

# 1) Estimate propensity scores with logistic regression
ps_model <- glm(
  d ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 +
      X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20,
  data = df_ipw,
  family = binomial(link = "logit")
)

df_ipw$ps <- predict(ps_model, type = "response")

# Overlap/common support diagnostic BEFORE trimming
ggplot(df_ipw, aes(x = ps, color = factor(d), fill = factor(d))) +
  geom_density(alpha = 0.25) +
  labs(
    title = "Propensity Score Overlap (Before Trimming)",
    x = "Estimated propensity score",
    y = "Density",
    color = "Treatment (d)",
    fill = "Treatment (d)"
  ) +
  theme_minimal()

# 2) Trim propensity scores to reduce extreme weights under weak overlap
df_ipw$ps_trim <- pmin(pmax(df_ipw$ps, 0.01), 0.99)

# Function requested in prompt: two-step IPW ATE with logistic PS + trimming
estimate_ipw_ate <- function(data, trim = 0.01) {
  ps_fit <- glm(
    d ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 +
      X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20,
    data = data,
    family = binomial(link = "logit")
  )

  ps_hat <- predict(ps_fit, newdata = data, type = "response")
  ps_hat <- pmin(pmax(ps_hat, trim), 1 - trim)

  mean(data$y * (data$d / ps_hat - (1 - data$d) / (1 - ps_hat)))
}

# 3) Manual IPW ATE using slide formula:
#    ATE = mean( y * ( d/ps - (1-d)/(1-ps) ) )
ate_ipw_manual <- estimate_ipw_ate(df_ipw, trim = 0.01)

cat("\nManual IPW ATE (primary estimate):", round(ate_ipw_manual, 4), "\n")

# 4) Bootstrap inference for the full two-step IPW estimator:
#    re-sample rows, re-estimate propensity model, re-calculate ATE each draw.
set.seed(123)
B_ipw <- 1000
n_ipw <- nrow(df_ipw)
boot_ipw_ate <- numeric(B_ipw)

for (b in seq_len(B_ipw)) {
  idx <- sample.int(n_ipw, size = n_ipw, replace = TRUE)
  boot_data <- df_ipw[idx, ]
  boot_ipw_ate[b] <- estimate_ipw_ate(boot_data, trim = 0.01)
}

ipw_boot_mean <- mean(boot_ipw_ate)
ipw_boot_se <- sd(boot_ipw_ate)
ipw_boot_ci <- quantile(boot_ipw_ate, probs = c(0.025, 0.975))

cat("Bootstrap mean (IPW):", round(ipw_boot_mean, 4), "\n")
cat("Bootstrap SE (IPW):", round(ipw_boot_se, 4), "\n")
cat("95% percentile CI (IPW): [", round(ipw_boot_ci[1], 4), ",", round(ipw_boot_ci[2], 4), "]\n")

# 5) Secondary comparison: IPW weighted regression-adjustment model
df_ipw$ipw_weight <- ifelse(df_ipw$d == 1,
                            1 / df_ipw$ps_trim,
                            1 / (1 - df_ipw$ps_trim))

cat("\nInverse Probability Weights Summary:\n")
summary(df_ipw$ipw_weight)

lm_ipw <- lm(y ~ d + X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8 + X9 + X10 + X11 + X12 + X13 + X14 + X15 + X16 + X17 + X18 + X19 + X20,
             data = df_ipw, weights = ipw_weight)

cat("\nSecondary comparison: weighted lm IPW-RA model\n")
print(summary(lm_ipw))
print(coeftest(lm_ipw, vcov. = vcovHC(lm_ipw, type = "HC3")))
```

### IPW Critical Discussion Points
- Overlap/common support: if many propensity scores are near 0 or 1, inverse weights become extreme and the estimator becomes unstable with high variance.
- Weighting logic: IPW up-weights observations with treatment status that is unlikely given their covariates, creating a pseudo-population with improved covariate balance.
- Efficiency: IPW is unbiased under conditional independence and overlap, but often has higher variance than AIPW.

## Augmented Inverse Probability Weighting (AIPW) - Manual Estimation
This section implements AIPW manually using the doubly robust score:

$$
ATE = \frac{1}{N}\sum_{i=1}^{N} \left( \frac{T_i}{\hat{p}(X_i)}(Y_i - \hat{\mu}_{T=1}(X_i)) + \hat{\mu}_{T=1}(X_i) \right)
- \frac{1}{N}\sum_{i=1}^{N} \left( \frac{(1-T_i)}{1-\hat{p}(X_i)}(Y_i - \hat{\mu}_{T=0}(X_i)) + \hat{\mu}_{T=0}(X_i) \right)
$$

```{r}
# User-requested structure:
# df            : data frame
# treatment     : binary treatment indicator (0/1)
# sales         : numeric outcome
# covariates    : character vector with covariate names

df <- casedata_task3
df$treatment <- df$d
df$sales <- df$y
covariates <- paste0("X", 1:20)

# Basic checks
stopifnot(all(c("treatment", "sales") %in% names(df)))
stopifnot(all(covariates %in% names(df)))
stopifnot(all(na.omit(unique(df$treatment)) %in% c(0, 1)))

# Keep complete cases for variables used in AIPW
aipw_vars <- c("treatment", "sales", covariates)
df_aipw <- df[complete.cases(df[, aipw_vars]), aipw_vars]

# AIPW function returning the sample ATE using the manual formula
estimate_aipw_ate <- function(data, covariate_names, trim = 0.01) {
  # 1) Selection model (propensity score): P(T=1|X)
  ps_formula <- as.formula(
    paste("treatment ~", paste(covariate_names, collapse = " + "))
  )
  ps_model <- glm(ps_formula, data = data, family = binomial(link = "logit"))
  p_hat <- predict(ps_model, newdata = data, type = "response")

  # Trimming avoids extreme inverse-probability weights when overlap is weak
  p_hat <- pmin(pmax(p_hat, trim), 1 - trim)

  # 2) Outcome models fitted separately by treatment arm
  mu1_formula <- as.formula(
    paste("sales ~", paste(covariate_names, collapse = " + "))
  )
  mu0_formula <- mu1_formula

  outcome_model_treated <- lm(mu1_formula, data = data[data$treatment == 1, ])
  outcome_model_control <- lm(mu0_formula, data = data[data$treatment == 0, ])

  # Predicted potential outcomes for every unit
  mu1_hat <- predict(outcome_model_treated, newdata = data)
  mu0_hat <- predict(outcome_model_control, newdata = data)

  # 3) Doubly robust correction terms:
  #    residuals (Y - mu_hat) are reweighted by inverse propensity scores.
  #    If either the PS model or the outcome models are correctly specified,
  #    the estimator remains consistent.
  score_treated <- (data$treatment / p_hat) * (data$sales - mu1_hat) + mu1_hat
  score_control <- ((1 - data$treatment) / (1 - p_hat)) * (data$sales - mu0_hat) + mu0_hat

  mean(score_treated) - mean(score_control)
}

# Point estimate
aipw_ate <- estimate_aipw_ate(df_aipw, covariates, trim = 0.01)
cat("Manual AIPW ATE estimate:", round(aipw_ate, 4), "\n")

# 4) Bootstrap inference for AIPW
set.seed(123)
B <- 1000
n <- nrow(df_aipw)
boot_ate <- numeric(B)

for (b in seq_len(B)) {
  idx <- sample.int(n, size = n, replace = TRUE)
  boot_data <- df_aipw[idx, ]
  boot_ate[b] <- estimate_aipw_ate(boot_data, covariates, trim = 0.01)
}

aipw_se <- sd(boot_ate)
aipw_ci <- quantile(boot_ate, probs = c(0.025, 0.975))
aipw_boot_mean <- mean(boot_ate)

cat("Bootstrap mean (AIPW):", round(aipw_boot_mean, 4), "\n")
cat("Bootstrap SE (AIPW):", round(aipw_se, 4), "\n")
cat("95% Bootstrap CI (AIPW): [", round(aipw_ci[1], 4), ",", round(aipw_ci[2], 4), "]\n")
```

### AIPW Critical Discussion Points
- Doubly robust consistency: AIPW is consistent if either the propensity score model or the outcome regression models are correctly specified.
- Efficiency: Compared with simple IPW, AIPW is typically more stable and can achieve the semiparametric efficiency bound under correct specification.
- Debiasing logic: The residual-weighted terms, $(Y - \hat{\mu})$, correct the regression predictions using inverse-probability weighting.
- Overlap/common support: Propensity scores should stay away from 0 and 1; weak overlap creates extreme weights and unstable estimates, so diagnostics and trimming are important.

# Double Machine Learning (DML): PLR with Cross-Fitting
We now estimate the ATE with DoubleML using the partially linear regression model:

$$
Y = \beta D + g(X) + \epsilon, \qquad D = m(X) + \xi
$$

Cross-fitting (sample splitting) is used to reduce overfitting bias in nuisance estimation, and the orthogonal score makes the final $\beta$ less sensitive to small nuisance-model errors.

```{r}
# Data setup for DML
x_cols <- paste0("X", 1:20)
dml_vars <- c("y", "d", x_cols)
dml_df <- casedata_task3[complete.cases(casedata_task3[, dml_vars]), dml_vars]

# Binary treatment check
stopifnot(all(unique(dml_df$d) %in% c(0, 1)))

# DoubleML data backend
dml_data <- DoubleMLData$new(
  data = dml_df,
  y_col = "y",
  d_cols = "d",
  x_cols = x_cols
)

# Tasks for out-of-sample nuisance performance
# Use only X covariates as predictors to match DML nuisance definitions.
task_y_data <- dml_df[, c("y", x_cols)]
task_y <- TaskRegr$new(id = "outcome_task", backend = task_y_data, target = "y")

task_d_data <- dml_df[, x_cols]
task_d_data$d_factor <- factor(dml_df$d, levels = c(0, 1))
task_d <- TaskClassif$new(
  id = "treatment_task",
  backend = task_d_data,
  target = "d_factor",
  positive = "1"
)

# Helper to construct requested learners with a fallback if IDs differ by install
get_lrn <- function(primary_id, fallback_id = NULL, ...) {
  if (primary_id %in% mlr_learners$keys()) {
    return(lrn(primary_id, ...))
  }
  if (!is.null(fallback_id) && fallback_id %in% mlr_learners$keys()) {
    message("Learner ", primary_id, " not found; using fallback ", fallback_id)
    return(lrn(fallback_id, ...))
  }
  stop("Required learner not found: ", primary_id,
       if (!is.null(fallback_id)) paste0(" (fallback also missing: ", fallback_id, ")") else "")
}

# Run one DML specification: nuisance metrics + PLR estimate
run_dml_spec <- function(spec_name, regr_id, classif_id,
                         regr_fallback = NULL, classif_fallback = NULL) {
  set.seed(123)

  # Outcome nuisance model g(X): evaluate out-of-sample RMSE
  learner_y <- get_lrn(regr_id, regr_fallback)
  rr_y <- resample(task_y, learner_y$clone(), rsmp("cv", folds = 5), store_models = FALSE)
  rmse_y <- rr_y$aggregate(msr("regr.rmse"))

  # Selection nuisance model m(X): evaluate out-of-sample log-loss and error
  learner_d_class <- get_lrn(classif_id, classif_fallback, predict_type = "prob")
  rr_d <- resample(task_d, learner_d_class$clone(), rsmp("cv", folds = 5), store_models = FALSE)
  logloss_d <- rr_d$aggregate(msr("classif.logloss"))
  ce_d <- rr_d$aggregate(msr("classif.ce"))

  # PLR in DoubleML uses regression nuisance learners for both l(X) and m(X)
  # (classification learner above is used for requested selection-model diagnostics).
  learner_m_regr <- get_lrn(regr_id, regr_fallback)
  dml_plr <- DoubleMLPLR$new(
    data = dml_data,
    ml_l = learner_y$clone(),
    ml_m = learner_m_regr$clone(),
    n_folds = 5
  )
  dml_plr$fit()

  data.frame(
    Model = spec_name,
    ATE_Beta = as.numeric(dml_plr$coef),
    Std_Error = as.numeric(dml_plr$se),
    RMSE_Y = as.numeric(rmse_y),
    LogLoss_D = as.numeric(logloss_d),
    Classif_Error_D = as.numeric(ce_d),
    stringsAsFactors = FALSE
  )
}

# Three requested learner families
dml_lasso <- run_dml_spec(
  spec_name = "Lasso",
  regr_id = "regr.lasso",
  classif_id = "classif.lasso",
  regr_fallback = "regr.cv_glmnet",
  classif_fallback = "classif.cv_glmnet"
)

dml_rf <- run_dml_spec(
  spec_name = "Random Forest",
  regr_id = "regr.ranger",
  classif_id = "classif.ranger"
)

dml_xgb <- run_dml_spec(
  spec_name = "XGBoost",
  regr_id = "regr.xgboost",
  classif_id = "classif.xgboost"
)

dml_results <- rbind(dml_lasso, dml_rf, dml_xgb)
dml_results$CI_Lower_95 <- dml_results$ATE_Beta - 1.96 * dml_results$Std_Error
dml_results$CI_Upper_95 <- dml_results$ATE_Beta + 1.96 * dml_results$Std_Error

cat("\n=== DML (PLR) Comparison Across Learners ===\n\n")
dml_results_to_print <- dml_results
num_cols <- sapply(dml_results_to_print, is.numeric)
dml_results_to_print[num_cols] <- lapply(dml_results_to_print[num_cols], round, 4)
print(dml_results_to_print, row.names = FALSE)

# "Well-fitting" model selection rule:
# prioritize low out-of-sample nuisance errors (RMSE_Y and LogLoss_D),
# because DML relies on high-quality nuisance prediction rather than interpretability.
scale01 <- function(x) {
  rng <- range(x)
  if (diff(rng) == 0) return(rep(0, length(x)))
  (x - rng[1]) / diff(rng)
}

dml_results$fit_score <- scale01(dml_results$RMSE_Y) + scale01(dml_results$LogLoss_D)
best_idx <- which.min(dml_results$fit_score)

cat("\nBest-fitting nuisance combination (lower RMSE + LogLoss):",
    dml_results$Model[best_idx], "\n")
```

### DML Discussion Points
- DML solves regularization bias and overfitting bias: sample splitting and cross-fitting reduce overfitting from flexible learners.
- Predictive performance of nuisance models is central: lower RMSE/log-loss means cleaner residualization (orthogonalization) for treatment-effect estimation.
- Compared with weighting estimators, PLR-style partialling out avoids direct division by propensity scores, which can be more stable under weak overlap.
- Even with ML-based nuisance models, causal validity still depends on conditional independence and excluding bad controls (e.g., colliders).


# Comparison of Treatment Effects Across Methods

```{r}
# Extract treatment effects and standard errors from all methods
ate_regression <- coef(model_adj)["d"]
se_regression <- sqrt(diag(vcovHC(model_adj, type = "HC3")))["d"]

att_no_replace <- coef(model_no_replace)["d"]
se_no_replace <- sqrt(diag(vcovHC(model_no_replace, type = "HC3")))["d"]

att_with_replace <- coef(model_with_replace)["d"]
se_with_replace <- sqrt(diag(vcovCL(model_with_replace, cluster = ~subclass)))["d"]

ate_euc <- coef(lm_euc)["d"]
se_euc <- sqrt(diag(vcovHC(lm_euc, type = "HC3")))["d"]

ate_ipw <- ate_ipw_manual
se_ipw <- ipw_boot_se

ate_ipw_lm <- coef(lm_ipw)["d"]
se_ipw_lm <- sqrt(diag(vcovHC(lm_ipw, type = "HC3")))["d"]

ate_aipw <- aipw_ate
se_aipw <- aipw_se

# Calculate 95% confidence intervals
ci_lower_reg <- ate_regression - 1.96 * se_regression
ci_upper_reg <- ate_regression + 1.96 * se_regression

ci_lower_no_rep <- att_no_replace - 1.96 * se_no_replace
ci_upper_no_rep <- att_no_replace + 1.96 * se_no_replace

ci_lower_rep <- att_with_replace - 1.96 * se_with_replace
ci_upper_rep <- att_with_replace + 1.96 * se_with_replace

ci_lower_euc <- ate_euc - 1.96 * se_euc
ci_upper_euc <- ate_euc + 1.96 * se_euc

ci_lower_ipw <- ipw_boot_ci[1]
ci_upper_ipw <- ipw_boot_ci[2]

ci_lower_ipw_lm <- ate_ipw_lm - 1.96 * se_ipw_lm
ci_upper_ipw_lm <- ate_ipw_lm + 1.96 * se_ipw_lm

ci_lower_aipw <- aipw_ci[1]
ci_upper_aipw <- aipw_ci[2]

# Create base comparison table for classical/weighting estimators
comparison_table_base <- data.frame(
  Method = c(
    "Regression Adjustment",
    "Matching (1:1 No Replacement)",
    "Matching (1:1 With Replacement)",
    "Euclidean Distance Matching",
    "Inverse Probability Weighting (IPW) - Manual Formula",
    "IPW Weighted Regression (Secondary)",
    "Augmented IPW (AIPW) - Manual Formula"
  ),
  Estimand = c(
    "ATE",
    "ATT",
    "ATT",
    "ATT",
    "ATE",
    "ATE",
    "ATE"
  ),
  Inference = c(
    "HC3",
    "HC3",
    "Clustered (subclass)",
    "HC3",
    "Bootstrap (1000)",
    "HC3",
    "Bootstrap (1000)"
  ),
  Estimate = c(
    round(ate_regression, 4),
    round(att_no_replace, 4),
    round(att_with_replace, 4),
    round(ate_euc, 4),
    round(ate_ipw, 4),
    round(ate_ipw_lm, 4),
    round(ate_aipw, 4)
  ),
  Std_Error = c(
    round(se_regression, 4),
    round(se_no_replace, 4),
    round(se_with_replace, 4),
    round(se_euc, 4),
    round(se_ipw, 4),
    round(se_ipw_lm, 4),
    round(se_aipw, 4)
  ),
  CI_Lower = c(
    round(ci_lower_reg, 4),
    round(ci_lower_no_rep, 4),
    round(ci_lower_rep, 4),
    round(ci_lower_euc, 4),
    round(ci_lower_ipw, 4),
    round(ci_lower_ipw_lm, 4),
    round(ci_lower_aipw, 4)
  ),
  CI_Upper = c(
    round(ci_upper_reg, 4),
    round(ci_upper_no_rep, 4),
    round(ci_upper_rep, 4),
    round(ci_upper_euc, 4),
    round(ci_upper_ipw, 4),
    round(ci_upper_ipw_lm, 4),
    round(ci_upper_aipw, 4)
  ),
  row.names = NULL
)

# Add DML learner-specific estimates to the same comparison table
dml_comparison <- data.frame(
  Method = paste0("DML-PLR (", dml_results$Model, ")"),
  Estimand = rep("ATE", nrow(dml_results)),
  Inference = rep("Orthogonal score SE (cross-fit)", nrow(dml_results)),
  Estimate = round(dml_results$ATE_Beta, 4),
  Std_Error = round(dml_results$Std_Error, 4),
  CI_Lower = round(dml_results$CI_Lower_95, 4),
  CI_Upper = round(dml_results$CI_Upper_95, 4),
  row.names = NULL
)

comparison_table <- rbind(comparison_table_base, dml_comparison)

cat("\n=== Summary: Treatment Effects Across All Methods (with 95% CIs) ===\n\n")
print(comparison_table, row.names = FALSE)

cat("\nNote: ATT rows are matching estimands; all other rows report ATE.\n")
cat("DML rows include learner-specific PLR estimates with 5-fold cross-fitting and orthogonal-score inference.\n")
```
