---
title: "Selection Bias"
author: "Jan-Michael Becker"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    self-contained-math: true
toc: true
---

```{r}
#| label: load necessary packages
#| include: false

require(sjPlot)
require(kableExtra)
```

# Introduction

In this document, we will discuss the concept of selection bias and conditional independence (unconfoundedness) through selection on observables.

Under the assumption of unconfoundedness (selection on observables), we assume that the treatment assignment $T_i$ is independent of the potential outcomes ${Y_i (1), Y_i (0)}$, after controlling for the covariates `X`. This conditional independence assumption formally implies that:

$$Y_i(1) - Y_i(0) \perp T_i | X_i $$

which means that the average treatment effect (ATE) can be consistently estimated using regression analysis as long as we control for the covariates `X1`, `X2`, and `X3`.

# Data Generating Process

We again generate a dataset of 10 000 observations using the following data generating process.

This code is similar to the earlier code, except that we replace the random treatment assignment `T` with a function $g(X_1, X_2, X_3)$ that depends on the covariates `X1`, `X2`, and `X3`. In addition, the potential outcomes also again depend on the covariates `X1`, `X2`, and `X3`. Thus, T is not independent of the potential outcomes.

```{r}
#| label: data-generation

# Set seed for reproducibility
set.seed(13579)

# Define the number of observations
n <- 10000

# Generate covariates
X1 <- rnorm(n, mean = 50, sd = 10)
X2 <- rpois(n, lambda = 5)
X3 <- sample(c("Yes", "No"), size = n, replace = TRUE)

X3 <- ifelse(X3 == "Yes", 1, 0)

# Generate treatment assignment
# Treatement assignment will be dependent on X1, X2, X3
# Create a linear combination of covariates to form a log-odds for T
log_odds_T <- -3.5 - 0.2 * X1 + 3 * X2 - 2 * X3

# Calculate the probability of T being 1 using the logistic function
prob_T_1 <- 1 / (1 + exp(-log_odds_T))

# Generate T as binary treatment assignment (0 or 1) based on the probabilities
T <- rbinom(n, 1, prob_T_1)

# Generate potential outcomes with covariates
E_y1 = 100
E_y0 = 80
Y1 <- E_y1 + 0.5*X1 + 2*X2 - 10*X3 + rnorm(n, mean = 0, sd = 5)
Y0 <- E_y0 + 0.5*X1 + 2*X2 - 10*X3 + rnorm(n, mean = 0, sd = 5)

# The true average treatment effect / causal effect
tau <- mean(Y1) - mean(Y0)

# Calculate the observed outcome based on treatment assignment
Y <- Y0 + (Y1-Y0)*T

# Create a data frame
data <- data.frame(T, Y, Y1, Y0, X1, X2, X3)
```

Let's first verify that T is not independent of the potential outcomes.

If the treatment assignments is independent of the individuals' potential outcomes we can assume that $E[Y_i(1)]=E[Y_i(1)|T_i=1]$

```{r}
#| label: verification1

# E[Y(1)]
mean(Y1)

# E[Y(1)|T=1]
mean(Y1[T==1])
```

This is not the case.

Likewise, if the treatment assignments is independent of the individuals' potential outcomes we can assume that $E[Y_i(0)] = E[Y_i(0)|T_i=0]$

```{r}
#| label: verification2

# E[Y(0)]
mean(Y0)

# E[Y(0)|T=0]
mean(Y0[T==0])
```

This is also not the case. Of course, in reality we will not know $E[Y_i(0)]$ and $E[Y_i(0)|T_i=0]$

# Regression without covariates

Now, we want to estimate the causal effect from the DGP. The average treatment effect $\tau$ is still set to 20. 

In this example, we do not have treatment effect heterogeneity (selection on returns) so the ATET should be about the same.

If we estimate the following regression equation: $Y_i = \alpha + \beta T_i + \epsilon_i$ [without]{.underline} covariates `X` we expect the treatment effect to be biased:

```{r}
#| label: regression
#| echo: true

# Run a linear regression on the data to estimate the treatment effect
model <- lm(Y ~ T, data = data)

# Print the summary of the regression output
tab_model(model)
```

```{r}
#| label: interpretation
#| echo: false
#| message: false

# Extract the regression coefficients
coef <- summary(model)$coefficients

# Extract the estimated treatment effect
tau_hat <- coef(model)["T"]
tau_confint <- confint(model, "T")

# Print the estimated treatment effect
cat("The estimated treatment effect is", round(tau_hat, 2), "[", round(tau_confint,2), "]", "\n")

# Print the 'true' treatment effect
cat("The true sample treatement effect is", round(tau,2), "\n")
```

What we estimate is the **observed difference in the outcome** that is a combination of the ATET and the selection bias (see lecture 3):

$E[Y_i |T_i=1] - E[Y_i|T_i=0] = E[Y_i(1)|T_i=1] - E[Y_i(0)|T_i=1] + E[Y_i(0)|T_i=1] - E[Y_i(0)|T_i=0]$

Observed difference in the outcome $E[Y|T=1] - E[Y|T=0]$:

```{r}
mean(Y[T==1]) - mean(Y[T==0])
```

Average treatment effect on the treated (ATET) $E[Y_i(1)|T_i=1] - E[Y_i(0)|T_i=1]$:

```{r}
mean(Y1[T==1]) - mean(Y0[T==1]) 
```

Selection bias: Difference in untreated outcomes between treated and untreated individuals $E[Y_i(0)|T_i=1] - E[Y_i(0)|T_i=0]$:

```{r}
mean(Y0[T==1]) - mean(Y0[T==0])
```

We can see from the selection bias that the treated individuals have on average much higher untreated outcomes than the untreated. Thus, it seems that we particularly select customers that would also spend more without treatment.

Again, remember that ATET here is supposed to be equal to ATE.

**Note:** Of course, we can only calculate the ATET and selection bias in this example, because we know both potential outcomes from the DGP. In reality, we would only observe `Y`.

# Including covariates into the regression

We now include the covariates and estimate the following regression equation: $Y_i = \alpha + \beta T_i + \gamma X_i + \epsilon_i$

```{r}
#| label: regression2
#| echo: true

# Run a linear regression on the data to estimate the treatment effect
model2 <- lm(Y ~ T + X1 + X2 + X3, data = data)

# Print the summary of the regression output
tab_model(model2)
```

```{r}
#| label: interpretation2
#| echo: false
#| message: false

# Extract the regression coefficients
coef <- summary(model2)$coefficients

# Extract the estimated treatment effect
tau_hat <- coef(model2)["T"]
tau_confint <- confint(model2, "T")

# Print the estimated treatment effect
cat("The estimated treatment effect is", round(tau_hat, 2), "[", round(tau_confint,2), "]", "\n")

# Print the 'true' treatment effect
cat("The true sample treatement effect is", round(tau,2), "\n")
```

We can see that the estimate of the average treatment effect (here ATET=ATE) is now close to the true value.

Thus, we have achieved unconfoundedness (or conditional independence), where the treatment assignment is as good as random conditional on the covariates. In our example, there are no unmeasured confounders on which T depends that are not included in the regression.

Of course, we could also easily create such a situation in which we ignore one of the `X` in the regression.

# Conclusion

In this document, we discussed the concept of unconfoundedness (selection on observables) and showed how we can get to an unbiased estimate of the ATE when including the variables into the regression equation as control variables (regression adjustment). We also illustrated how we could calculate the selection bias when we would know the potential outcomes.
