---
title: "GRA4158 Marketing - Analysis of Experiments, Portfolio 4"
format: 
  html:
    embed-resources: true
    self-contained-math: true
toc: true
warning: false
author: Vilijam Cekov, Luiz Felipe De Souza Simao
---
```{r}
#| label: load necessary packages
#| include: false

library(synthdid)
library(dplyr)
library(readr)
library(plm)
library(lmtest)
library(knitr)
```

# Import Data
```{r}
#| label: Importing and processing the data
df <- read_delim("Store_data_2026_taks4.csv", delim = ";")

# Create necessary dummy variables
df <- df %>%
  mutate(
    # 'post_treatment' is 1 for the weeks after the marketing strategy started
    post_treatment = ifelse(Weekind >= 78, 1, 0),
    # 'treated' is the actual interaction / DiD term (1 ONLY for Store 109 AND Weekind >= 78)
    treated = ifelse(storeNum == 109 & Weekind >= 78, 1, 0)
  )

# Display the structure of the data to verify column names
str(df)

# Check the first few rows
head(df)
```
<!-- # Use only treatment + appropriate pre-treatment covariates -->

# A simple pre-post analysis of only the treated store for each product controlling for relevant covariates
The relevant covariates chosen are price and promotion for each product, as they are likely to influence sales. Time-invariant covariates such as store size and city size are excluded from this analysis since we are only looking at one store (Store 109) and they do not vary over time.

```{r}
#| label: pre-post analysis
# Filter data to only include Store 109
df_109 <- df %>% filter(storeNum == 109)

# --- Product 1: Pre-Post Model ---
pre_post_p1 <- lm(p1sales ~ post_treatment + p1price + p1prom, data = df_109)
summary(pre_post_p1)

# --- Product 2: Pre-Post Model ---
pre_post_p2 <- lm(p2sales ~ post_treatment + p2price + p2prom, data = df_109)
summary(pre_post_p2)

```

# A simple two-way fixed-effect panel model for each product controlling for relevant covariates
Store fixed effects will automatically absorb (control for) all time-invariant store characteristics (like storesize and city). Time fixed effects will absorb aggregate shocks that affect all stores in a given week. Therefor no particular changed will be made to the data, as the plm package will handle this automatically when we specify the model.

```{r}
#| label: two-way fixed effects analysis
pdf <- pdata.frame(df, index = c("storeNum", "Weekind"))

# --- Product 1: TWFE Model ---
twfe_p1 <- plm(p1sales ~ treated + p1price + p1prom, 
               data = pdf, 
               model = "within", 
               effect = "twoways")
summary(twfe_p1)

# --- Product 2: TWFE Model ---
twfe_p2 <- plm(p2sales ~ treated + p2price + p2prom, 
               data = pdf, 
               model = "within", 
               effect = "twoways")
summary(twfe_p2)
```

Correcting for autocorrelation and heteroskedasticity

```{r}
#| label: Extracting Robust Standard Errors for TWFE
# --- Product 1 TWFE with Clustered Standard Errors ---
coeftest(twfe_p1, vcov = vcovHC(twfe_p1, type = "HC1", cluster = "group"))

# --- Product 2 TWFE with Clustered Standard Errors ---
coeftest(twfe_p2, vcov = vcovHC(twfe_p2, type = "HC1", cluster = "group"))
```

# Simple DiD, SC, and SDID analyses for each product
### Product 1: p1sales
```{r}
#| label: DiD, SC, and SDID analyses for Product 1

# Ensure data frame is standard class (synthdid prefers plain data.frames over tibbles)
df <- as.data.frame(df)

# Setup panel matrices for Product 1
setup_p1 <- panel.matrices(
  df,
  unit = "storeNum",
  time = "Weekind",
  outcome = "p1sales",
  treatment = "treated"
)

# Estimate the treatment effects for Product 1
sdid_est_p1 <- synthdid_estimate(setup_p1$Y, setup_p1$N0, setup_p1$T0)
did_est_p1 <- did_estimate(setup_p1$Y, setup_p1$N0, setup_p1$T0)
sc_est_p1 <- sc_estimate(setup_p1$Y, setup_p1$N0, setup_p1$T0)
estimates_p1 = list(did_est_p1, sc_est_p1, sdid_est_p1)

# Print the estimates
names(estimates_p1) = c('Diff-in-Diff', 'Synthetic Control', 'Synthetic Diff-in-Diff')
print(unlist(estimates_p1))

# Calculate standard errors and 95% Confidence Intervals using placebo method for SDID
se_p1 <- sqrt(vcov(sdid_est_p1, method = "placebo"))
cat(sprintf("(SDID) 95%% CI: [%.2f, %.2f]\n", sdid_est_p1 - 1.96 * se_p1, sdid_est_p1 + 1.96 * se_p1))

# Calculate standard errors and 95% Confidence Intervals using placebo method for DID
se_p1 <- sqrt(vcov(did_est_p1, method = "placebo"))
cat(sprintf("(DID) 95%% CI: [%.2f, %.2f]\n", did_est_p1 - 1.96 * se_p1, did_est_p1 + 1.96 * se_p1))

# Calculate standard errors and 95% Confidence Intervals using placebo method for SC
se_p1 <- sqrt(vcov(sc_est_p1, method = "placebo"))
cat(sprintf("(SC) 95%% CI: [%.2f, %.2f]\n", sc_est_p1 - 1.96 * se_p1, sc_est_p1 + 1.96 * se_p1))

# Plot 1: Trends of treated vs. synthetic control
synthdid_plot(estimates_p1, facet.vertical = FALSE)
# Plot 2: See which control stores received the highest weights
synthdid_units_plot(estimates_p1)
```

### Product 2: p2sales
The same steps as for Product 1 are repeated for Product 2.
```{r}
#| label: DiD, SC, and SDID analyses for Product 2
# Setup panel matrices for Product 2
setup_p2 <- panel.matrices(
  df,
  unit = "storeNum",
  time = "Weekind",
  outcome = "p2sales",
  treatment = "treated"
)

# Estimate the treatment effects for Product 2
sdid_est_p2 <- synthdid_estimate(setup_p2$Y, setup_p2$N0, setup_p2$T0)
did_est_p2 <- did_estimate(setup_p2$Y, setup_p2$N0, setup_p2$T0)
sc_est_p2 <- sc_estimate(setup_p2$Y, setup_p2$N0, setup_p2$T0)
estimates_p2 = list(did_est_p2, sc_est_p2, sdid_est_p2)

# Print the estimates
names(estimates_p2) = c('Diff-in-Diff', 'Synthetic Control', 'Synthetic Diff-in-Diff')
print(unlist(estimates_p2))

# Calculate standard errors and 95% Confidence Intervals using placebo method for SDID
se_p2 <- sqrt(vcov(sdid_est_p2, method = "placebo"))
cat(sprintf("(SDID) 95%% CI: [%.2f, %.2f]\n", sdid_est_p2 - 1.96 * se_p2, sdid_est_p2 + 1.96 * se_p2))

# Calculate standard errors and 95% Confidence Intervals using placebo method for DID
se_p2 <- sqrt(vcov(did_est_p2, method = "placebo"))
cat(sprintf("(DID) 95%% CI: [%.2f, %.2f]\n", did_est_p2 - 1.96 * se_p2, did_est_p2 + 1.96 * se_p2))

# Calculate standard errors and 95% Confidence Intervals using placebo method for SC
se_p2 <- sqrt(vcov(sc_est_p2, method = "placebo"))
cat(sprintf("(SC) 95%% CI: [%.2f, %.2f]\n", sc_est_p2 - 1.96 * se_p2, sc_est_p2 + 1.96 * se_p2))

# Plot 1: Trends of treated vs. synthetic control
synthdid_plot(estimates_p2, facet.vertical = FALSE)
# Plot 2: See which control stores received the highest weights
synthdid_units_plot(estimates_p2)
```

## Regressing out the covariates
Given that the base synthdid package used does not easily accept additional control variables (like price and promotions) directly into its primary matrix setup, we can implement a solution based on the Frisch-Waugh-Lovell (FWL) theorem: we first use a linear regression to predict sales based only on the covariates, extract the residuals (the portion of sales unexplained by price, promotions, etc.), and then apply the SDID model to those residuals.

By adding the overall mean back to the residuals, we keep the interpretation on the same scale as the original sales volume.

### Product 1: p1sales
```{r}
#| label: Regressing out covariates for Product 1

# Run a linear model and extract the residuals (unexplained sales)
model_p1 <- lm(p1sales ~ p1price + p1prom + compind + storesize + citysize, data = df)
df$p1sales_adj <- residuals(model_p1) + mean(df$p1sales, na.rm = TRUE)

# Setup panel matrices using the NEW adjusted outcome variable
setup_p1_adj <- panel.matrices(
  df,
  unit = "storeNum",
  time = "Weekind",
  outcome = "p1sales_adj", 
  treatment = "treated"
)

# Estimate the adjusted SDID Treatment Effect
sdid_est_p1_adj <- synthdid_estimate(setup_p1_adj$Y, setup_p1_adj$N0, setup_p1_adj$T0)
print(sdid_est_p1_adj)

# Calculate and print confidence intervals for Product 1
se_p1_adj <- sqrt(vcov(sdid_est_p1_adj, method = "placebo"))
cat(sprintf("Adjusted 95%% CI: [%.2f, %.2f]\n\n", 
            sdid_est_p1_adj - 1.96 * se_p1_adj, 
            sdid_est_p1_adj + 1.96 * se_p1_adj))

# Visualize adjusted trends
plot(sdid_est_p1_adj)+theme_minimal()
```

### Product 2: p2sales

```{r}
#| label: Regressing out covariates for Product 2

# Run a linear model and extract the residuals (unexplained sales)
model_p2 <- lm(p2sales ~ p2price + p2prom + compind + storesize + citysize, data = df)
df$p2sales_adj <- residuals(model_p2) + mean(df$p2sales, na.rm = TRUE)

# Setup panel matrices using the NEW adjusted outcome variable
setup_p2_adj <- panel.matrices(
  df,
  unit = "storeNum",
  time = "Weekind",
  outcome = "p2sales_adj", 
  treatment = "treated"
)

# Estimate the adjusted SDID Treatment Effect
sdid_est_p2_adj <- synthdid_estimate(setup_p2_adj$Y, setup_p2_adj$N0, setup_p2_adj$T0)
print(sdid_est_p2_adj)

# Calculate and print confidence intervals for Product 2
se_p2_adj <- sqrt(vcov(sdid_est_p2_adj, method = "placebo"))
cat(sprintf("Adjusted 95%% CI: [%.2f, %.2f]\n\n", 
            sdid_est_p2_adj - 1.96 * se_p2_adj, 
            sdid_est_p2_adj + 1.96 * se_p2_adj))

# Visualize adjusted trends
plot(sdid_est_p2_adj)+theme_minimal()
```

# Summary of Findings

```{r}
#| label: compile-results-table
#| echo: false
#| message: false

# 1. Extract robust standard errors from the TWFE models
# (We run coeftest again here to easily extract the values)
robust_twfe_p1 <- coeftest(twfe_p1, vcov = vcovHC(twfe_p1, type = "HC1", cluster = "group"))
robust_twfe_p2 <- coeftest(twfe_p2, vcov = vcovHC(twfe_p2, type = "HC1", cluster = "group"))

# 2. Create a consolidated data frame with Estimates and Standard Errors (SE)
results_summary <- data.frame(
  Model = c(
    "Pre-Post (LM)", 
    "Two-Way Fixed Effects (TWFE)", 
    "Difference-in-Differences (DID)", 
    "Synthetic Control (SC)", 
    "Synthetic DID (SDID)", 
    "Adjusted SDID (with Covariates)"
  ),
  
  # Product 1 Estimates and Standard Errors
  P1_Estimate = c(
    coef(pre_post_p1)["post_treatment"],
    robust_twfe_p1["treated", "Estimate"],
    as.numeric(did_est_p1),
    as.numeric(sc_est_p1),
    as.numeric(sdid_est_p1),
    as.numeric(sdid_est_p1_adj)
  ),
  P1_SE = c(
    summary(pre_post_p1)$coefficients["post_treatment", "Std. Error"],
    robust_twfe_p1["treated", "Std. Error"],
    as.numeric(sqrt(vcov(did_est_p1, method = "placebo"))),
    as.numeric(sqrt(vcov(sc_est_p1, method = "placebo"))),
    as.numeric(sqrt(vcov(sdid_est_p1, method = "placebo"))),
    as.numeric(sqrt(vcov(sdid_est_p1_adj, method = "placebo")))
  ),
  
  # Product 2 Estimates and Standard Errors
  P2_Estimate = c(
    coef(pre_post_p2)["post_treatment"],
    robust_twfe_p2["treated", "Estimate"],
    as.numeric(did_est_p2),
    as.numeric(sc_est_p2),
    as.numeric(sdid_est_p2),
    as.numeric(sdid_est_p2_adj)
  ),
  P2_SE = c(
    summary(pre_post_p2)$coefficients["post_treatment", "Std. Error"],
    robust_twfe_p2["treated", "Std. Error"],
    as.numeric(sqrt(vcov(did_est_p2, method = "placebo"))),
    as.numeric(sqrt(vcov(sc_est_p2, method = "placebo"))),
    as.numeric(sqrt(vcov(sdid_est_p2, method = "placebo"))),
    as.numeric(sqrt(vcov(sdid_est_p2_adj, method = "placebo")))
  )
)

# 3. Generate the HTML table
kable(
  results_summary, 
  digits = 2, 
  col.names = c("Model", "P1 Estimate", "P1 SE", "P2 Estimate", "P2 SE"),
  caption = "Comparison of Treatment Effect Estimates for Product 1 and Product 2",
  align = "lcccc"
)
```